Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-17 docs/proposals/NSF/ProjDesc/ProjDesc.tex
working on restrucuturing and writing

%---------------------------------------------------------------------------------------------------%
%% INTRODUCTION
\section{Proposed Research}

\subsection{Introduction and Background}

Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water.  Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats.  Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can compromise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the UN report on water resources for further detail on the threats to water resources~\cite{UNwater}.
% Could also include USGS if needed.

@@

This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium, for example the anomalous dispersion of tracers. A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $\mathcal{O}(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis. Dynamic erosion of porous media is a highly active area of research, as closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments a great opportunity for comparison against our numerical simulations. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute, which would help to guide, calibrate, and validate the computations (see letter of collaboration).

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. Color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. 
}
%The same simulation as above with velocity magnitude in color. Velocity magnitude highlights the most dominant channels whose growth is reinforced by the shape-flow interaction of erosion.
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing occur over much shorter timescales.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the project proposed here is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models~\cite{Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.

%---------------------------------------------------------------------------------------------------%
\subsection{Governing equations}

We first summarize governing equations underlying the proposed work.
At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations
\begin{equation}
\label{StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, and $\mu$ the fluid viscosity. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations in Eq.~\eqref{StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Eq.~\eqref{StokesEq} is a microscopic description is required to study the shape evolution of individual grains due to fine-scale erosive effects, either mechanical or chemical. The description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

While the most immediate proposed work on computational methods to simulate erosion will utilize~\eqref{StokesEq}, the later stages of this project will also incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models~\cite{Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

%---------------------------------------------------------------------------------------------------%
\subsection{Mechanical and chemical erosion of a porous medium}

The first problem discussed here is the erosion of a porous medium, as has been preliminarily investigated in~\cite{Quaife2018}. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, erosion results from surface shear stress
\begin{align}
\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{StokesEq}. 
The motion of the interface is proportional to the absolute value of shear
\begin{align}
\Vn = \CE \, \abs{\tau},
\end{align}
In the latter case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
\label{Ceq}
& \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C \, ,
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \left. \Diff \pd{C}{\ny} \right. %\right\vert_{\bdry} \, ,
\end{equation}
Both processes have been investigated by the PIs and collaborators in
the context of single bodies in high Reynolds-number
flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In
the low-Reynolds regime of porous-media flow, numerical methods have
been developed by the PIs to simulate shear-induced
erosion~\cite{Quaife2018}. These simulations combine highly accurate
boundary integral methods with stable interface evolution techniques.
The boundary integral solver is accelerated by the fast multipole
method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$
complexity. Meanwhile, the so-called {\thL} method is used for interface
evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.
%A near-term goal is to adapt these methods to handle the case of chemically-induced dissolution.

%\subsubsection{Basic dynamics}

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
%Erosion not only diminishes the size of the bodies but also alters their shapes considerably. A few well-defined channels develop in the space between bodies. Several of the bodies vanish in finite time and the simulation continues without interruption.
}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Fig.~\ref{80circ8vort} shows preliminary results of multiple bodies undergoing mechanical erosion in Stokes flow. The figure shows 80 bodies of initially random size and position, undergoing erosion. This figure extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015}, which enables closer contact between bodies and thus more bodies.
The color shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity. Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. The channelization process is made more evident by visualization of the velocity magnitude, as seen in Fig.~\ref{80circ8vel}.

%\subsubsection{Evolution of macroscopic properties}

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows new, preliminary results on how media properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies erode. 
%Time is normalized by the value $t_f$ at which all bodies vanish and hence the area fraction is zero. 
Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more mild decrease in area fraction is not sufficient to account for this exponential rate, and hence the reshaping process, in particular the formation of channels, plays a pivotal role.

Lastly, the right-most plot of Fig.~\ref{fig3} shows the anisotropy, which is calculated as the ratio of the vertical resistivity to the horizontal resistivity, see Eq.~\eqref{DarcyEq}. The vertical resistivity is computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertically than horizontally.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
\caption{\label{fig3} Left: the area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$. Middle: the resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: the anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.
}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

%\subsubsection{Immediate goals of proposed work}

An immediate goal of the proposed work is extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. Near-term plans are to include flow transport and gravitational sedimentation of the eroding grains. For this extension, contact forces between bodies must be included to eliminate unphysical contact that is inevitable due to discretization errors.

>>>>

%---------------------------------------------------------------------------------------------------%
%% INTRODUCTION
\section{Introduction and Background}
Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water.  Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats.  Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can compromise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the UN report on water resources for further detail on the threats to water resources~\cite{UNwater}.
% Could also include USGS if needed.

@@

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution. {\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(100-1000)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} Dynamic erosion of porous media is a highly active area of research, as closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments a great opportunity for comparison against our numerical simulations. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing occur over much shorter timescales.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models~\cite{Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.

%----------------------------------------------------------------------%
\subsection{Governing Equations}
At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations \begin{equation}
\label{eqn:StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the
fluid viscosity, and $\Omega$ is the fluid domain. Linear elliptic PDEs,
such as the Stokes equations, can be reformulated as a boundary integral
equation (BIE).  BIEs have several numerical advantages including a
dimension reduction, automatic mass conservation, and high-order
accuracy. An equivalent BIE formulation of the Stokes equations in
Eq.~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Eq.~\eqref{eqn:StokesEq}, and its equivalent integral equation formulation, are microscopic descriptions that are required to study the shape evolution of individual grains due to fine-scale erosive effects, either mechanical or chemical. The description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

While the most immediate proposed work on computational methods to
simulate erosion will utilize~\eqref{eqn:StokesEq}, the later stages of this project will also incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models~\cite{Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

%----------------------------------------------------------------------%
\subsection{Project Overview}

\paragraph{Broader Impacts:} The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The high-fidelity, efficient computational methods proposed here will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

\paragraph{Intellectual Merit:} The proposed research introduces a host of new computational challenges and opportunities. First the range of scales is vast: spatial scales range from microscopic granular constituents to large geological aquifers; timescales range from that of a sudden sinkhole collapse to years required mechanical and chemical erosion. The systems are inherently multicomponent, with coupling between the fluid and solid phases. Although the governing PDEs are linear, the presence of moving boundaries introduces nonlinear feedback between geometry and flow. To tackle these challenges, the PIs will combine cutting-edge computational tools with reduced-order modeling. Mixed-scale, deep neural networks will be used to learn from the data generated by high-fidelity numerical simulations to parameterize coarse-grained models based on the multiphase framework. Additionally, controlled laboratory experiments will be used to guide and verify theory developed herein.

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Computational plays an essential role to simulate erosion of porous medium.

  \item The proposed research is directly applicable to several geophysical applications.

  \item The proposed research is a collaborative effort and will include graduate, undergraduate, and K--12 students
\end{itemize}

\paragraph{Results from Prior NSF Support:} Neither PI has previous funding from NSF.

\paragraph{PIs' Qualifications:} PI Moore joined the Department of Mathematics at Florida State University (FSU) in 2014, and is also a Research Associate in FSU's Geophysical Fluid Dynamics Institute (GFDI).  His expertise includes modeling dynamic interactions between fluids and structures, especially in geophysical and biologically motivated settings. PI Quaife joined the Department of Scientific Computing at FSU in 2015 and is also a Research Associate in GFDI. His expertise includes high-fidelity and efficient numerical methods for fluid dynamic in complex geometries.

The PIs co-authored a paper describing numerical methods for simulating erosion~\cite{Quaife2018} in a viscous fluid. Very recently, a second paper has been submitted to the Journal of Fluid Mechanics~\cite{chi-moo-qua2019}, where additional numerical methods are developed and characterizations of transport are presented.

\paragraph{Outline of the Proposal:}

\todo[inline]{Summary of what is in the proposal.}

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
\todo[inline]{Discuss what the main research questions are here}
\begin{itemize}[noitemsep]
  \item Characterize anisotropy and permeability to inform homogenized
    models including Darcy's relation.

  \item Improve numerical methods including FMM, preconditioners, and
    quadrature.

  \item Simulate erosion via chemical reactions ie.~dissolution.

  \item Simulate transport of finite Peclet number species in eroded
    geometries.
\end{itemize}

%----------------------------------------------------------------------%
\subsection{Mechanical and Chemical Erosion of a Porous Medium}
With the development of efficient and high-fidelity numerical methods to solve the Stokes equations, we are able to simulate erosion. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Figs.~\ref{80circ8vort} and~\ref{80circ8vel} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. The color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows new, preliminary results on how media properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies erode. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more mild decrease in area fraction is not sufficient to account for this exponential rate, and hence the reshaping process, in particular the formation of channels, plays a pivotal role.

Lastly, the right-most plot of Fig.~\ref{fig3} shows the anisotropy, which is calculated as the ratio of the vertical resistivity to the horizontal resistivity. The vertical resistivity is computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} {\em Left}: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$. {\em Middle}: The resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. However, the implementation of the quadrature method does not fully use FMM acceleration. The numerical challenge is developing the required FMM expansions for the velocity, vorticity, and deformation tensor layer potentials. Since we require several layer potentials with different kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and its implementations that are readily available (for example, see \url{http://pvfmm.org}). Other challenges that will be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to allow for stable simulations to user-specified tolerances. PI Quaife has experience developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.
